Physics of Tidal Dissipation in Early-Type Stars and White Dwarfs: Hydrodynamical Simulations of Internal Gravity Wave Breaking in Stellar Envelopes

### 2020-04-13 writeups/Su_IGW_break.tex
filter out

We aim to model wave dynamics over multiple density and pressure scaleheights.
Start with the compressible Euler equations.
\begin{align}
    \partial_t \rho + \p{\bm{u}\cdot\bm{\nabla}} \rho
            + \rho \bm{\nabla}\cdot\bm{u} =\, &0, \\
    \partial_t s + \p{\bm{u}\cdot\bm{\nabla}} s =\, &0, \\
    \partial_t \bm{u} + \p{\bm{u}\cdot\bm{\nabla}}\bm{u} +
        \frac{1}{\rho}\bm{\nabla} p =& \, \bm{g},
\end{align}
where $s$ is the entropy. The pressure is calculated from the entropy and
density using the equation of state
\begin{equation}
    \frac{s}{c_p} = \log \frac{p^{1/\gamma}}{\rho},
\end{equation}
where $c_p$ is the specific heat at constant pressure, and $\gamma$ is the ratio
of specific heats. For computational ease, it is convenient to filter the fast
sound waves from these equations. One approach is to assume pressure
perturbations are small \citep[yielding the ``pseudo-incompressible''
equations,][]{anel_part2}, or that all thermodynamic perturbations are small
\citep[yielding the ``anelastic'' equations,][]{anel_part1}. In these
approximations, one of the thermodynamic equations is replaced by a constraint
equation: $\bm{\nabla}\cdot \p{p_0^{1/\gamma} \bm{u}}=0$ for
pseudo-incompressible; $\bm{\nabla}\cdot\p{\rho_0 \bm{u}} =0$ for anelastic.
Here $p_0$ and $\rho_0$ are the background density and pressure profiles. The
pressure in the momentum equation can be interpreted as a Lagrange multiplier
which enforces the constraint \citep{anel_part2}. Upon linearization, both
approximations conserve a wave energy
\begin{equation}
    E_w = \frac{1}{2}\rho_0 \abs{\bm{u}}^2
            + \frac{1}{2}\frac{g^2}{N^2} \frac{(\rho')^2}{\rho_0}
        = \frac{1}{2}\rho_0 \abs{\bm{u}}^2
            + \frac{1}{2}\rho_0 N^2 \frac{(s')^2}{\abs{\bm{\nabla} s_0}^2},
            \label{eq:ewave}
\end{equation}
where $\rho'$ and $s'$ represent the density and entropy perturbations.

Rather than assume thermodynamic perturbations are small, we instead filter
sound waves by taking the limit $\gamma\rightarrow \infty$. Then the entropy and
log density are proportional to each other, and the entropy equation becomes
\begin{equation}
    \partial_t \log \rho + \p{\bm{u}\cdot \bm{\nabla}} \log \rho = 0.
\end{equation}
Together with mass conservation, this implies
\begin{equation}
    \bm{\nabla} \cdot \bm{u} = 0.
\end{equation}
We solve these equations together with the normal momentum equation.

>>>>

We aim to model wave dynamics over multiple density and pressure scaleheights.
Start with the compressible Euler equations.
\begin{align}
    \partial_t \rho + \p{\bm{u}\cdot\bm{\nabla}} \rho
            + \rho \bm{\nabla}\cdot\bm{u} =\, &0, \\
    \partial_t s + \p{\bm{u}\cdot\bm{\nabla}} s =\, &0, \\
    \partial_t \bm{u} + \p{\bm{u}\cdot\bm{\nabla}}\bm{u} +
        \frac{1}{\rho}\bm{\nabla} p =& \, \bm{g},
\end{align}
where $s$ is the entropy. The pressure is calculated from the entropy and
density using the equation of state
\begin{equation}
    \frac{s}{c_p} = \log \frac{p^{1/\gamma}}{\rho},
\end{equation}
where $c_p$ is the specific heat at constant pressure, and $\gamma$ is the ratio
of specific heats. For computational ease, it is convenient to filter out the
fast sound waves from these equations. One approach is to assume pressure
perturbations are small \citep[yielding the ``pseudo-incompressible''
equations,][]{anel_part2}, or that all thermodynamic perturbations are small
\citep[yielding the ``anelastic'' equations,][]{anel_part1}. In these
approximations, one of the thermodynamic equations is replaced by a constraint
equation: $\bm{\nabla}\cdot \p{p_0^{1/\gamma} \bm{u}}=0$ for
pseudo-incompressible; $\bm{\nabla}\cdot\p{\rho_0 \bm{u}} =0$ for anelastic.
Here $p_0$ and $\rho_0$ are the background density and pressure profiles. The
pressure in the momentum equation can be interpreted as a Lagrange multiplier
which enforces the constraint \citep{anel_part2}. Upon linearization, both
approximations conserve a wave energy
\begin{equation}
    E_w = \frac{1}{2}\rho_0 \abs{\bm{u}}^2
            + \frac{1}{2}\frac{g^2}{N^2} \frac{(\rho')^2}{\rho_0}
        = \frac{1}{2}\rho_0 \abs{\bm{u}}^2
            + \frac{1}{2}\rho_0 N^2 \frac{(s')^2}{\abs{\bm{\nabla} s_0}^2},
            \label{eq:ewave}
\end{equation}
where $\rho'$ and $s'$ represent the density and entropy perturbations.

Rather than assume thermodynamic perturbations are small, we instead filter out
sound waves by taking the limit $\gamma\rightarrow \infty$. Then the entropy and
log density are proportional to each other, and the entropy equation becomes
\begin{equation}
    \partial_t \log \rho + \p{\bm{u}\cdot \bm{\nabla}} \log \rho = 0.
\end{equation}
Together with mass conservation, this implies
\begin{equation}
    \bm{\nabla} \cdot \bm{u} = 0.
\end{equation}
We solve these equations together with the normal momentum equation.
